High-level cognition is supported by at least second order dynamic correlations in neural activity patterns

### 2019-08-23 paper/boneyard.tex
working on methods-- dynamic correlations

>>>>

A major challenge to studying such patterns is that typically neither
the correlations nor the hierarchical organizations of those
correlations may be directly observed.  Rather, these fundamental
properties must be inferred indirectly by examining the observable
parts of the system-- e.g., the behaviors of the individual units of
that system.  Here we propose a series of mathematical operations that may be used to approximate dynamic correlations at a range of scales (i.e., orders of interaction).

There are two basic steps to our approach (Fig.~\ref{fig:methods_fig}).  In the first step, we take
a number-of-timepoints ($T$) by number-of-features ($F$)
\textit{matrix of observations} ($\mathbf{X}$) and we return a $T$ by
$\frac{F^2 - F}{2}$ \textit{matrix of dynamic correlations}
($\mathbf{Y}$).  Here $\mathbf{Y_0}$ describes, at each moment, how
all of the features (columns of $\mathbf{X}$) are inferred to be
interacting.  (Since the interactions are assumed to be non-recurrent
and symmetric, only the upper triangle of the full correlation matrix
is computed.)  In the second step, we project $\mathbf{Y_0}$ onto an
$F$-dimensional space, resulting in a new $T$ by $F$ matrix
$\mathbf{Y_1}$.  Note that $\mathbf{Y_1}$ contains information about
the correlation dynamics present in $\mathbf{X}$, but represented in a
compressed number of dimensions.  By repeatedly applying these two
steps in sequence, we can examine and explore higher order dynamic
correlations in $\mathbf{X}$.

### 2019-08-23 paper/main.tex
working on methods-- dynamic correlations

We validated our approach using synthetic data where the underlying
correlations were known.  We then applied our framework to a
neuroimaging dataset collected as 125 participants listened to either
an audio recording of a ten-minute story or a temporally scrambled
version of the story, or underwent a resting state
scan~\citep{SimoEtal16}.  We used a subset of the data to train
across-participant classifiers to decode listening times using a blend
of neural features (comprising neural activity patterns, as well as
different orders of correlations between those patterns that were
inferred using our computational framework).  We found that both the
PCA-based and eigenvector centrality-based approaches yielded neural
patterns that could be used to decode accurately.  Both approaches
also yielded the best decoding accuracy for data collected during
(intact) story listening when high-order (PCA:
second-order; eigenvector centrality: fourth-order) dynamic
correlation patterns were included as features.  When we trained
classifiers on the scambled stories or resting state data, only
lower-order dynamic patterns were informative to the decoders.  Taken
together, our results indicate that high-level cognition is supported
by high-order dynamic patterns of communication between brain
structures.  \textbf{JRM STOPPED HERE}

\section*{Methods}
A major challenge to studying such patterns is that typically neither
the correlations nor the hierarchical organizations of those
correlations may be directly observed.  Rather, these fundamental
properties must be inferred indirectly by examining the observable
parts of the system-- e.g., the behaviors of the individual units of
that system.  Here we propose a series of mathematical operations that may be used to approximate dynamic correlations at a range of scales (i.e., orders of interaction).

There are two basic steps to our approach (Fig.~\ref{fig:methods_fig}).  In the first step, we take
a number-of-timepoints ($T$) by number-of-features ($F$)
\textit{matrix of observations} ($\mathbf{X}$) and we return a $T$ by
$\frac{F^2 - F}{2}$ \textit{matrix of dynamic correlations}
($\mathbf{Y}$).  Here $\mathbf{Y_0}$ describes, at each moment, how
all of the features (columns of $\mathbf{X}$) are inferred to be
interacting.  (Since the interactions are assumed to be non-recurrent
and symmetric, only the upper triangle of the full correlation matrix
is computed.)  In the second step, we project $\mathbf{Y_0}$ onto an
$F$-dimensional space, resulting in a new $T$ by $F$ matrix
$\mathbf{Y_1}$.  Note that $\mathbf{Y_1}$ contains information about
the correlation dynamics present in $\mathbf{X}$, but represented in a
compressed number of dimensions.  By repeatedly applying these two
steps in sequence, we can examine and explore higher order dynamic
correlations in $\mathbf{X}$.

\subsection*{Dynamic correlations}
Given a matrix of observations, we can compute the (static)
correlations between any pair of observations, $\mathbf{X}_i$ and
$\mathbf{X}_j$ using:
\begin{align}
  \mathrm{corr}(\mathbf{X}_i, \mathbf{X}_j) &= \frac{\sum_{t=1}^T \left(\mathbf{X}_i(t)
                                              -
                                              \bar{\mathbf{X}_i}\right) \left(\mathbf{X}_j(t)
                                              -
                                              \bar{\mathbf{X}_j}\right)}{\sqrt{\sum_{t=1}^T
                                              \sigma^2_{\mathbf{X}_i} 
                                              \sigma^2_{\mathbf{X}_j}}},~\mathrm{where}\\\label{eqn:corr}
  \bar{\mathbf{X}_k} &= \sum_{t=1}^T
                       \mathbf{X}_k(t),~\mathrm{and}\\
  \sigma^2_{\mathbf{X}_k} &= \sum_{t=1}^T \left( \mathbf{X}_k -
                            \bar{\mathbf{X}_k} \right)^2 
\end{align}

We can generalize this formula to compute time-varying correlations by
incorporating a \textit{weight function} that takes a time $t$ as input,
and returns how much the observed data every timepoint (including $t$)
contribute to the correlations at time $t$ (Fig.~\ref{fig:weights}).
\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{figs/kernels}
  \caption{\textbf{Examples of time-varying weights.  Each panel
      displays per-timepoint weights at $t = 50$, evaluated for 100
      timepoints ($1, ..., 100$).  a. Uniform weights.} The weights
    are timepoint-invariant; observations at all timepoints are
    weighted equally, and do not change as a function of $t$.  This is
    a special case of weight function that reduces dynamic
    correlations to static correlations.  \textbf{b. Dirac delta
      function.} Only the observation at timepoint $t$ is given weight
    (of 1), and weights for observations at all other timepoints are
    set to 0.  \textbf{c. Gaussian weights.} Each observation's
    weights fall off in time according to a Gaussian probability
    density function centered on $\mu = t$.  Weights derived using
    several different example variance parameters ($\sigma^2$ are
    displayed.  \textbf{d. Laplace weights.}  Each observation's
    weights fall off in time according to a Laplace probability
    density function centered on $\mu = t$.  Weights derived using
    several different example scale parameters ($b$) are displayed.
    \textbf{e. Mexican hat (Ricker wavelet) weights.}  Each
    observation's weights fall off in time according to a Ricker
    wavelet centered on $t$.  This function highlights the
    \textit{contrasts} between local versus surrounding activity
    patterns in estimating dynamic correlations. Weights derived using
    several different example width parameters ($\sigma$) are
    displayed.  }
  \label{fig:weights}
\end{figure}

>>>>

We validated our approach using synthetic data where the underlying
correlations were known.  We then applied our framework to a
neuroimaging dataset collected as 125 participants listened to either
an audio recording of a ten-minute story or a temporally scrambled
version of the story, or underwent a resting state
scan~\citep{SimoEtal16}.  We used a subset of the data to train
across-participant classifiers to decode listening times using a blend
of neural features (comprising neural activity patterns, as well as
different orders of correlations between those patterns that were
inferred using our computational framework).  We found that both the
PCA-based and eigenvector centrality-based approaches yielded neural
patterns that could be used to decode accurately.  Both approaches
also yielded the best decoding accuracy for data collected during
(intact) story listening when high-order (PCA:
second-order; eigenvector centrality: fourth-order) dynamic
correlation patterns were included as features.  When we trained
classifiers on the scambled stories or resting state data, only
lower-order dynamic patterns were informative to the decoders.  Taken
together, our results indicate that high-level cognition is supported
by high-order dynamic patterns of communication between brain
structures.

\section*{Methods}
Our general approach to comprises four general steps.  First, we
derive a kernel-based approach to computing dynamic pairwise
correlations in a $T$ (timepoints) by $K$ (features) multivariate
timeseries, $\mathbf{X}_0$.  This yields a $T$ by $\mathcal{O}(K^2)$ matrix of
dynamic correlations, $\mathbf{Y}_1$, where each row comprises the upper
triangle of the correlation matrix at a single timepoint, reshaped
into a row vector.  Second, we apply a dimensionality reduction step
to project the matrix of dynamic correlations back onto a
$K$-dimensional space.  This yields a $T$ by $K$ matrix, $\mathbf{X}_1$, that
reflects an approximation of the dynamic correlations reflected in the
original data.  Third, we use repeated applications of the
kernel-based dynamic correlation step to $\mathbf{X}_n$ and the dimensionality
reduction step to the resulting $\mathbf{Y}_{n+1}$ to estimate high-order
dynamnic correlations.  Each application of these steps to a $T$ by
$K$ time series $\mathbf{X}_n$ yields a $T$ by $K$ matrix, $\mathbf{X}_{n+1}$, that
reflects the dynamic correlations between the columns of $\mathbf{X}_n$.  In
this way, we refer to $n$ as the \textit{order} of the timeseries,
where $\mathbf{X}_0$ (order 0) denotes the original data and $\mathbf{X}_n$ denotes
$n^\mathrm{th}$-order dynamic correlations between the columns of
$\mathbf{X}_0$.  Finally, we use a cross-validation--based decoding approach to
evaluate how well information contained in a given order (or weighted
mixture of orders) may be used to decode relevant cognitive states.
If including a given $\mathbf{X}_n$ in the feature set yields higher
classification accuracy on held-out data, we interpret this as
evidence that the given cognitive states are reflected in patterns of
$n^\mathrm{th}$-order correlations.

\subsection*{Kernel-based approach for computing dynamic correlations}
Given a matrix of observations, we can compute the (static)
correlations between any pair of columns, $\mathbf{X}(\cdot, i)$ and
$\mathbf{X}(\cdot, j)$ using:
\begin{align}
  \mathrm{corr}(\mathbf{X}(\cdot, i), \mathbf{X}(\cdot, j)) &=
                                                              \frac{\sum_{t=1}^T
                                                              \left(\mathbf{X}(t,
                                                              i)
                                                              -
                                                              \bar{\mathbf{X}(\cdot,
                                                              i)}\right)
                                                              \left(\mathbf{X}(t,
                                                              j)
                                                              -
                                                              \bar{\mathbf{X}(\cdot, j)}\right)}{\sqrt{\sum_{t=1}^T
                                                              \sigma^2_{\mathbf{X}(\cdot, i)} 
                                                              \sigma^2_{\mathbf{X}(\cdot, j)}}},~\mathrm{where}\\\label{eqn:corr}
  \bar{\mathbf{X}(\cdot, k)} &= \sum_{t=1}^T
                       \mathbf{X}(t, k),~\mathrm{and}\\
  \sigma^2_{\mathbf{X}(\cdot, k)} &= \sum_{t=1}^T \left( \mathbf{X}(t, k) -
                            \bar{\mathbf{X}(\cdot, k)} \right)^2 
\end{align}

\textbf{JRM STOPPED HERE}
We can generalize this formula to compute time-varying correlations by
incorporating a \textit{weight function} that takes a time $t$ as input,
and returns how much the observed data every timepoint (including $t$)
contribute to the correlations at time $t$ (Fig.~\ref{fig:weights}).
\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{figs/kernels}
  \caption{\textbf{Examples of time-varying weights.  Each panel
      displays per-timepoint weights at $t = 50$, evaluated for 100
      timepoints ($1, ..., 100$).  a. Uniform weights.} The weights
    are timepoint-invariant; observations at all timepoints are
    weighted equally, and do not change as a function of $t$.  This is
    a special case of weight function that reduces dynamic
    correlations to static correlations.  \textbf{b. Dirac delta
      function.} Only the observation at timepoint $t$ is given weight
    (of 1), and weights for observations at all other timepoints are
    set to 0.  \textbf{c. Gaussian weights.} Each observation's
    weights fall off in time according to a Gaussian probability
    density function centered on $\mu = t$.  Weights derived using
    several different example variance parameters ($\sigma^2$ are
    displayed.  \textbf{d. Laplace weights.}  Each observation's
    weights fall off in time according to a Laplace probability
    density function centered on $\mu = t$.  Weights derived using
    several different example scale parameters ($b$) are displayed.
    \textbf{e. Mexican hat (Ricker wavelet) weights.}  Each
    observation's weights fall off in time according to a Ricker
    wavelet centered on $t$.  This function highlights the
    \textit{contrasts} between local versus surrounding activity
    patterns in estimating dynamic correlations. Weights derived using
    several different example width parameters ($\sigma$) are
    displayed.  }
  \label{fig:weights}
\end{figure}
